---
title: Matrix Decomposition with Kronecker Weights
author: 
    - name: Jan de Leeuw
      orcid: 0000-0003-1420-1797
      email: jan@deleeuwpdx.net
      affiliation: 
        - name: University of California Los Angeles
          city: Los Angeles
          state: CA
          url: www.ucla.edu
      license: "CC0"
date: last-modified
date-format: long
bibliography: [mypubs.bib, total.bib]
number-sections: true
pdf-engine: lualatex
keep-tex: true
format:
   pdf:
    fontsize: 12pt
    include-in-header: 
     - preamble.tex
    keep-tex: true
    link-citations: true
    documentclass: scrartcl
    number-sections: true
   html:
    fontsize: 12pt
    include-in-header: 
     - preamble.css
    keep-md: true
    number-sections: true
toc: true
toc-depth: 3
editor: source
papersize: letter
graphics: true
link-citations: true
mainfont: Times New Roman
abstract: The familiar orthogonal decomposition of a matrix into row effect, column effects, and interactions is generalized to a matrix space with the Kronecker inner product.
---

```{r loadpackages, echo = FALSE}
suppressPackageStartupMessages(library(knitr, quietly = TRUE))
suppressPackageStartupMessages(library(tinytex, quietly = TRUE))
```

```{r echo = FALSE}
source("decompose.R")
source("checker.R")
```

\sectionbreak 

**Note:** This is a working manuscript which will be expanded/updated
frequently. All suggestions for improvement are welcome. All Rmd, tex,
html, pdf, R, and C files are in the public domain. Attribution will be
appreciated, but is not required. The files can be found at
<https://github.com/deleeuw/decomp> 

\sectionbreak

# Kronecker Weights

Suppose $R$ and $C$ are positive semi-definite matrices of order $n$ and $m$.
They define an inner product on the space $\mathbb{R}^{n\times m}$ of 
$n\times m$ matrices as
$\langle X, Y\rangle_{RC}=\text{tr}\ RXCY'$ and a corresponding squared
norm (pseudo-norm if either $R$ or $C$ is singular) as
$\|X\|_{RC}^2=\langle X, X\rangle_{RC}$. 

Two important alternative representations of this inner product are 
\begin{equation}
\langle X,Y\rangle_{RC} = \text{vec}(X)'(C\otimes R)\text{vec}(Y),
\end{equation}
or, elementwise,
\begin{equation}
\langle X,Y\rangle_{RC} = \sum_{i=1}^{n}\sum_{j=1}^{m}\sum_{k=1}^{n}\sum_{\ell=1}^{m} r_{ik}c_{j\ell}x_{ij}y_{k\ell}.
\end{equation}

Since we will only be taking about inner products of this form we will leave out the subscript $RC$ from now on.

# Matrix Decomposition

## Subspaces

We use $e_n$ and $e_m$ for vectors with $n$ and $m$ elements equal to one. If the number of elements is clear from the context we simply use $e$. 

Define four subspaces of $\mathbb{R}^{n\times m}$. 

1. $\mathbb{L}_\mu$, constant matrices, of the form $\mu e_ne_m'$.
2. $\mathbb{L}_a$, matrices with all columns the same, of the form $ae_m'$.
3. $\mathbb{L}_b$, matrices with all rows the same, of the form $e_nb'$.
4. $\mathbb{L}_h$, matrices orthogonal to the previous three spaces.

We will choose $a, b, \mu$ and $H\in\mathbb{L}_h$ such that these four subspacs are orthogonal
in the Kronecker inner product. For this we need
\begin{subequations}
\begin{align}
\text{tr}\ R\mu ee'Cea'&=\mu a'Re\times e'Ce=0,\\
\text{tr}\ R\mu ee'Cbe'&=\mu b'Ce\times e'Re=0,\\
\text{tr}\ Rae'Cbe'&=a'Re\times b'Ce=0,\\
\text{tr}\ R\mu ee'CH'&=\mu e'CH'Re=0,\\
\text{tr}\ Rae'CH'&=e'CH'Ra=0,\\
\text{tr}\ Reb'CH'&=b'CH'Re=0.
\end{align}
\end{subequations}

It follows the four subspaces are orthogonal if $a'Re=b'Ce=0$ and
$HCe$ and $H'Re$ are zero. Thus we redefine our subspaces as

1. $\mathbb{L}_\mu$: all matrices $\mu ee'$.
2. $\mathbb{L}_a$: all matrices $ae'$ where $a'Re=0$.
3. $\mathbb{L}_b$: all matrices $eb'$ where $b'Ce=0$.
4. $\mathbb{L}_h$: all $H$ with $HCe=0$ and $H'Re=0$.

## Projection

Now project, in the Kronecker metric, a matrix $X$ on each of these orthogonal subspaces.

### First

First project on $\mathbb{L}_\mu$. We must minimize 
\begin{equation}
\|X-\mu ee'\|^2=\|X\|^2-2\mu e'RXCe+\mu^2e'Re\times e'Ce,
\end{equation}
over $\mu$, with solution
\begin{equation}
\hat\mu=\frac{e'RXCe}{e'Re\times e'Ce}.\label{eq-mu}
\end{equation}

### Second

Now project on $\mathbb{L}_a$. We have
\begin{equation}
\|X-ae'\|^2=\|X\|^2-2a'RXCe+a'Ra\times e'Ce,
\end{equation}
which must be minimized over all $a'Re=0$. With a Lagrange multiplier $\lambda$
the stationary equations are
\begin{equation}
RXCe-\lambda Re=e'Ce\times Ra.
\end{equation}
Premultiply by $e'$ gives
\begin{equation}
\hat\lambda=\frac{e'RXCe}{e'Re},
\end{equation}
and thus 
\begin{equation}
e'Ce\times\hat a=XCe-e\frac{e'RXCe}{e'Re},\label{eq-asolve}
\end{equation}
or
\begin{equation}
\hat a=\left(I-\frac{ee'R}{e'Re}\right)X\left(\frac{Ce}{e'Ce}\right).
\end{equation}

At this point it is convenient to define the projectors
\begin{align}
P:=\frac{ee'R}{e'Re},\label{eq-pdef}\\
Q:=\frac{Cee'}{e'Ce}.\label{eq-qdef}
\end{align}
The projection on $\mathbb{L}_a$ is 
\begin{equation}
\hat ae'=(I-P)XQ.
\end{equation}
In addition, from \eqref{eq-mu}, 
\begin{equation}
\hat\mu ee'=PXQ.
\end{equation}

### Third

Projecting on $\mathbb{L}_b$ is similar to projecting on $\mathbb{L}_a$, by symmetry. Thus
\begin{equation}
e\hat b'=PX(I-Q).
\end{equation}

### Fourth

If $P$ and $Q$ are arbitrary square matrices of orders $n$ and $m$, not necessarily projectors, we have
\begin{equation}
X=PXQ+(I-P)XQ+PX(I-Q)+(I-P)X(I-Q).
\end{equation}
For our decomposition it follows that if we define the "residual"
\begin{equation}
\hat H:=X-\hat\mu ee'-\hat ae'-e\hat b'\label{eq-hdef},
\end{equation}
then $\hat H=(I-P)X(I-Q)$.
Because $H$ in \eqref{eq-hdef} is the residual after projecting on a subspace
the orthogonality of the residual space
to the row and column spaces is basically Pythagoras' Theorem. 
Alternatively, however, we could
have defined $\hat H$ directly as the projection of $X$ on $\mathbb{L}_h$.
This produces the same $\hat H$. Here follows a proof of this last statement. 

The stationary equations for the projection, using two vectors $\phi$ and $\psi$ of Lagrange multipliers, are
\begin{subequations}
\begin{align}
H&=X-\phi e'-e\psi',\label{eq-hstat1}\\
HCe &= 0,\label{eq-hstat2}\\
H'Re&=0.\label{eq-hstat3}
\end{align}
\end{subequations}
Postmultiply \eqref{eq-hstat1} by $Ce$. Then we must have
$0=XCe-\phi e'Ce-e\psi'Ce$ and thus 
\begin{subequations}
\begin{equation}
\phi=\frac{XCe}{e'Ce}-\frac{ee'C}{e'Ce}\psi.\label{eq-phicalc}
\end{equation}
In the same way
\begin{equation}
\psi=\frac{X'Re}{e'Re}-\frac{ee'R}{e'Re}\phi.\label{eq-psicalc}
\end{equation}
\end{subequations}
Substitute \eqref{eq-psicalc} into \eqref{eq-phicalc}. Then
\begin{equation}
\phi=\frac{XCe}{e'Ce}-\frac{ee'C}{e'Ce}\left\{\frac{X'Re}{e'Re}-\frac{ee'R}{e'Re}\phi\right\}.
\end{equation}
Collecting terms and using \eqref{eq-asolve} and \eqref{eq-pdef} gives $(I-P)\phi=\hat a$. In the same way $(I-Q)\psi=\hat b$. 
This implies that $\phi=\hat a+\alpha e$ and $\psi=\hat b+\beta e$ for some scalars $\alpha$ and $\beta$, so that
\begin{equation}
\hat H=X-\hat ae'-e\hat b'-(\alpha+\beta)ee'.\label{eq-hcalc}
\end{equation}
Premultiplying with $e'R$ and postmultiplying with $Ce$, and then using
\eqref{eq-hstat2} and \eqref{eq-hstat3}, gives that $\alpha+\beta=\hat\mu$.
End of proof.

Define $\mathbb{L}_h$ as the subspace of all matrices of this form, or equivalently the matrices with $HCe=0$ and $e'RH=0$.

Because $(I-Q)Ce=0$ it follows that
\begin{subequations}
\begin{equation}
\text{tr}\ R\hat HC(\hat\mu ee')=0
\end{equation}
and 
\begin{equation}
\text{tr}\ R\hat HC(e\hat b')=0
\end{equation}
and
\begin{equation}
\text{tr}\ R\hat ae'C\hat H'=0
\end{equation}
\end{subequations}
Thus $\mathbb{L}_h$ is also orthogonal in the Kronecker inner product to the other three subspaces. 

## Pythagoras

It follows that in the Kronecker norm
\begin{equation}
\|X\|^2=\|PXQ\|^2+\|(I-P)XQ\|^2+\|PX(I-Q)\|^2+\|(I-P)X(I-Q)\|^2,\label{eq-decomp}
\end{equation}
and thus we not only have a decomposition of the data $X$, but also of its squared Kronecker norm. 

This is important when fitting $Y$ to $X$, with various restrictions on $Y$.
Suppose, for example, we require $Y=\mu ee'+ae'+eb'+H$, with the orthogonality or identifiability constraints $a'Re=b'Ce=0$ and $HCe=0$ and $H'Re=0$, and possibly with additional separate constraints on $\mu$, $a$, $b$, and $H$. We may require, for example, that
$\text{rank}(H)\leq p$, with $p<\min(m-1,n-1)$ and that $a$ and $b$ are
increasing. We compute the projections first and then fit $Y$. From
\eqref{eq-decomp}
We can write \eqref{eq-decomp} as
\begin{multline}
\|X-Y\|^2=(\hat\mu-\mu)^2(e'Re)^2(e'Ce)^2+\\\{(\hat a-a)'R(\hat a-a)\}^2(e'Ce)^2+\{(\hat b-b)'C(\hat b-b)\}^2(e'Re)^2+\|(\hat H-H)\|^2.
\end{multline}
we can minimize each of the four components separately. 

In the example with monotonicity constraints on $a$ and $b$ and rank constraints\
on $H$
that means using monotone regression to minimize $(\hat a-a)'R(\hat a-a)$
and $(\hat b-b)'C(\hat b-b)$ over increasing $a$ and $b$ and minimizing
$\text{tr} R(\hat H-H)C(\hat H-H)'$ with $H=AB'$ over the $n\times p$ matrices
$A$ and the $m\times p$ matrices $B$. For this last problem the stationary equations for $A$ and $B$ are (ignoring the orthogonality constraints
of $H$ for the time being)
\begin{subequations}
\begin{align}
R\hat HCB&=RA(B'CB),\\
C\hat H'RA&=CB(A'RA).
\end{align}
\end{subequations}
If we define $\tilde B:=C^\frac12 B$ and $\tilde A:=R^\frac12 A$ then
\begin{subequations}
\begin{align}
\left\{R^{-\frac12}\hat HC^{-\frac12}\right\}\tilde B&=\tilde A(\tilde B'\tilde B),\\
\left\{C^{-\frac12}\hat H'R^{-\frac12}\right\}\tilde A&=\tilde B(\tilde A'\tilde A),
\end{align}
\end{subequations}
showing that $\tilde A$ and $\tilde B$ can be computed from the
singular value decomposition of $R^{-\frac12}\hat HC^{-\frac12}$.
Since $\tilde HCe=0$ we see that $C^\frac12e$ is a singular vector
with singular value zero, and the other singular vectors in
$\tilde B$ can be chosen orthogonal to $C^\frac12e$. Thus
$\tilde B'C^\frac12e=B'Ce=0$, and because $H=AB'$ we have $HCe=0$.
In the same way $H'Re=0$, and the orthogonality/identifiability conditions on
$H$ are satisfied.


## Special Case

The classical special case of our results has $R=I$ and $C=I$. Then
$P$ and $Q$ are symmetric projectors, with $P=ee'/n$ and $Q=ee'/m$.
Vector $\hat a$ are the centered row means, $\hat b$ are the centered
column means, $\hat\mu$ is the grand mean, and matrix $\hat H$ is the
doubly-centered $X$. Orthogonality of the subspaces forms the
basis for Factorial Analysis of Variance.

## Singular R and/or C

If either $R$ or $C$ is singular then some of the analysis in the previous 
sections must be modified. 

For the decomposition
\begin{equation}
X=\mu ee'+ae'+eb'+H
\end{equation}
consider the case with $Ce=0$. The projections on $\mathbb{L}_\mu$ and
$\mathbb{L}_a$ are both zero, so that we are basically looking at
the projection on $\mathbb{L}_b$ only, with no need to require
orthogonality. The projection is $e\hat b'=PX$ and thus 
the residual is $\hat H=(I-P)X$. If both $Ce=0$ and $Re=0$ all
three projections are zero and $\hat H=X$.

A singularity $Cz=0$ for some $z\neq e$ is less serious.

# Generalizations

The developments in this paper can be generalized in several ways. 

## Regression

The vectors $e_n$ and $e_m$ in our formulas can be replaced by any known vectors $u$ and $v$ with $n$ and $m$ elements. In fact they can
be replaced by known $n\times r$ and $m\times r$ matrices $U$ and $V$ of rank $r$. 
Assume without loss of generality that $U'RU=V'CV=I$.
 
### Subspaces 

1. $\mathbb{L}_M$: $UMV'$ with $M$ of dimension $r\times r$, 
2. $\mathbb{L}_A$: $UA$ with $A$ of dimension $r\times m$, 
3. $\mathbb{L}_B$: $B'V'$ with $B$ of dimension $r\times n$, 
4. $\mathbb{L}_H$: the residual space. 

Thus the decomposition we are looking for is
\begin{equation}
X=UMV'+UA+B'V'+H,
\end{equation}
which is a form of regressing $X$ on $U$ and $V$. Or, alternatively,
the residuals $H$ are $X$ "corrected for" $U$ and $V$.

### Orthogonality

1. $\mathbb{L}_M\perp\mathbb{L}_A$ if $\text{tr}\ U'RUMV'CA'=0$ thus if $ACV=0$
2. $\mathbb{L}_M\perp\mathbb{L}_B$ if $\text{tr}\ BRUMV'CV=0$ thus if $BRU=0$
3. $ACV=0$ and $BRU=0$ already imply that $\mathbb{L}_A\perp\mathbb{L}_B$. 

### Projection

We define the two projectors $P:=UU'R$ and $Q:=CVV'$.

1. Projection on $\mathbb{L}_M$ using $\hat M=U'RXCV$ is $PXQ$.
2. Projection on $\mathbb{L}_A$ using $\hat A=U'RX(I-Q)$ is $PX(I-Q)$.
3. Projection on $\mathbb{L}_B$ using $\hat B'=(I-P)XCV$ is $(I-P)XQ$.
4. Projection on $\mathbb{L}_H$ is $(I-P)X(I-Q)$.

## Multidimensional Arrays

The same reasoning applies to multidimensional arrays. If the array has $p$
"dimensions" then there are $2^p$ components. The squared Kronecker inner product does not have a simple matrix representation like $\text{tr}\ RXCX'$ any more. But, in case of $n_1\times n_2\times n_3$ arrays $X$ and $Y$ with positive semi-definite weight matrices $R$, $C$, and $S$ of orders $n_1, n_2$ and $n_3$, we still have
\begin{equation}
\langle X,Y\rangle = \text{vec}(X)'(S\otimes C\otimes R)\text{vec}(Y)
\end{equation}
or, elementwise,
\begin{equation}
\langle X,Y\rangle = \sum_{i=1}^{n_1}\sum_{j=1}^{n_2}\sum_{k=1}^{n_3}\sum_{i'=1}^{n_1}\sum_{j'=1}^{n_2}\sum_{k'=1}^{n_3} r_{ii'}c_{jj'}s_{kk'}x_{ijk}y_{i'j'k'}
\end{equation}
The eight subspaces are of the form
$x_{ijk}=\mu,
x_{ijk}=a_i,
x_{ijk}=b_j,
x_{ijk}=t_j,
x_{ijk}=\gamma_{ij},
x_{ijk}=\delta_{ik},
x_{ijk}=\epsilon_{jk},$
and a final "residual" space. The orthogonality of the subspaces
and the formulas for the projections can be worked out in the same
way as for the two-dimensional case. I may do this in some other paper,

## Look Ma, No Coordinates

The results apply to any finite sequence of orthogonal subspaces in any finite-dimensional inner product space. Suppose $\Pi_1,..,\Pi_t$ are projectors (linear idempotent operators) on orthogonal $\mathbb{L}_1,\cdots,\mathbb{L}_t$,
and $\circ$ is composition. Thus $\Pi_s=\Pi_s\circ\Pi_s$ for all $s$ and $\Pi_s\circ \Pi_t=0$ for all $s\neq t$. If $\Pi_r=I-(\Pi_1+\cdots+\Pi_t)$ then $\Pi_s\circ \Pi_r=0$ for all $s$. Thus the residual space is orthogonal to all $\mathbb{L}_s$. The decomposition is
\begin{equation}
X=\Pi_1(X)+\Pi_2(X)+\cdots+\Pi_r(X).
\end{equation}
In our examples from the previous sections we have $\Pi_1(X)=PXQ$, $\Pi_2(X)=PX(I-Q)$,
$\Pi_3(X)=(I-P)XQ$, and $\Pi_4(X)=(I-P)X(I-Q)$.


\sectionbreak


# Example

```{r}
set.seed(12345)
xmat <- matrix(rnorm(12), 4, 3)
rmat <- crossprod(matrix(rnorm(40), 10, 4)) / 10
cmat <- crossprod(matrix(rnorm(30), 10, 3)) / 10
```
```{r echo = FALSE}
print("xmat")
print(xmat)
print("cmat")
print(cmat)
print("rmat")
print(rmat)
```

\sectionbreak

We now apply the decompose() function from the code section.

```{r echo}
h<-decompose(xmat, rmat, cmat)
```
The four components are:
```{r echo = FALSE}
print(h$dd)
```

\sectionbreak

The Kronecker inner products of the four components are:
```{r echo = FALSE}
print(h$ip)
```
Check: the trace of the inner product matrix and the squared norm of the data:
```{r echo = FALSE}
print(h$ssq)
```

\sectionbreak

# Code

```{r eval = FALSE}
set.seed(12345)
xmat <- matrix(rnorm(12), 4, 3)
rmat <- crossprod(matrix(rnorm(40), 10, 4)) / 10
cmat <- crossprod(matrix(rnorm(30), 10, 3)) / 10

decompose <- function(xmat, rmat, cmat) {
  n <- nrow(xmat)
  m <- ncol(xmat)
  pmat <- matrix(colSums(rmat), n, n, byrow = TRUE) / sum(rmat)
  qmat <- matrix(rowSums(cmat), m, m) / sum(cmat)
  px <- pmat %*% xmat
  xq <- xmat %*% qmat
  dd <- rep(list(0),4) 
  dd[[1]] <- px %*% qmat
  dd[[2]] <- px - dd[[1]]
  dd[[3]] <- xq - dd[[1]]
  dd[[4]] <- xmat - dd[[1]] - dd[[2]] - dd[[3]]
  ip <- matrix(0, 4, 4)
  for (i in 1:4) {
    for (j in 1:4) {
      ip[i, j] <- kroneckerIP(dd[[i]], dd[[j]], rmat, cmat)
    }
  }
  ssq <- c(kroneckerIP(xmat, xmat, rmat, cmat), sum(diag(ip)))
  return(list(dd = dd, ip = ip, ssq = ssq))
}

kroneckerIP <- function(xmat, ymat, rmat, cmat) {
  return(sum(ymat * (rmat %*% xmat %*% cmat)))
}
```

# References

