---
title: Matrix Decomposition with Kronecker Weights
author: 
    - name: Jan de Leeuw
      orcid: 0000-0003-1420-1797
      email: jan@deleeuwpdx.net
      affiliation: 
        - name: University of California Los Angeles
          city: Los Angeles
          state: CA
          url: www.ucla.edu
      license: "CC0"
date: last-modified
date-format: long
bibliography: [mypubs.bib, total.bib]
number-sections: true
pdf-engine: lualatex
keep-tex: true
format:
   pdf:
    fontsize: 12pt
    include-in-header: 
     - preamble.tex
    keep-tex: true
    link-citations: true
    documentclass: scrartcl
    number-sections: true
   html:
    fontsize: 12pt
    include-in-header: 
     - preamble.css
    keep-md: true
    number-sections: true
toc: true
toc-depth: 3
editor: source
papersize: letter
graphics: true
link-citations: true
mainfont: Times New Roman
abstract: The familiar orthogonal decomposition of a matrix into row effect, column effects, and interactions is generalized to a matrix space with the Kronecker inner product.
---

```{r loadpackages, echo = FALSE}
suppressPackageStartupMessages(library(knitr, quietly = TRUE))
suppressPackageStartupMessages(library(tinytex, quietly = TRUE))
```

```{r echo = FALSE}
source("decompose.R")
source("checker.R")
```

\sectionbreak 

**Note:** This is a working manuscript which will be expanded/updated
frequently. All suggestions for improvement are welcome. All Rmd, tex,
html, pdf, R, and C files are in the public domain. Attribution will be
appreciated, but is not required. The files can be found at
<https://github.com/deleeuw> 

\sectionbreak

# Kronecker Weights

$R$ and $C$ are positive semidefinite matrices of order $n$ and $m$.
They define an inner product on the space $\mathbb{R}^{n\times m}$ of 
$n\times m$ matrices as
$\langle X, Y\rangle_{RC}=\text{tr}\ RXCY'$ and a corresponding squared
norm (pseudo-norm if either $R$ or $C$ is singular) as
$\|X\|_{RC}^2=\langle X, X\rangle_{RC}$. From now on we leave out the
subscript $RC$.

# Matrix Decomposition

Define three subspaces of $\mathbb{R}^{n\times m}$. We use $e$ for vectors with all elements equal to one. The number of elements in $e$ will be clear from the context, otherwise we will use $e_m$ and $e_n$.

1. $\mathbb{L}_\mu$, constant matrices, $X=\mu e_ne_m'$.
2. $\mathbb{L}_a$, matrices with all columns the same, $X=ae_m'$.
3. $\mathbb{L}_b$, matrices with all rows the same $X=e_nb'$.

We will hoose $a, b, \mu$ such that these three subspacs are orthogonal
in the Kronecker inner product. For this we need
\begin{subequations}
\begin{align}
\text{tr}\ R\mu ee'Cea'&=\mu a'Re\times e'Ce=0,\\
\text{tr}\ R\mu ee'Cbe'&=\mu b'Ce\times e'Re,\\
\text{tr}\ Rae'Cbe'&=a'Re\times b'Ce.
\end{align}
\end{subequations}

It follows the three subspaces are orthogonal if $a'Re=b'Ce=0$. Thus we
redefine our subspaces as

1. $\mathbb{L}_\mu$: all $X$ with $X=\mu ee'$.
2. $\mathbb{L}_a$: all $X$ with $X=ae'$ where $a'Re=0$.
3. $\mathbb{L}_b$: all $X$  with $X=eb'$ where $b'Ce=0$.

Now project, in the Kronecker metric, matrix $X$ on each of these orthogonal subspaces.

First on $\mathbb{L}_\mu$. Minimize over $\mu$
\begin{equation}
\|X-\mu ee'\|^2=\|X\|^2-2\mu e'RXCe+\mu^2e'Re\times e'Ce,
\end{equation}
and thus
\begin{equation}
\hat\mu=\frac{e'RXCe}{e'Re\times e'Ce}.\label{eq-mu}
\end{equation}
Now over $\mathbb{L}_a$. We have
\begin{equation}
\|X-ae'\|^2=\|X\|^2-2a'RXCe+a'Ra\times e'Ce,
\end{equation}
which must be minimized over all $a'Re=0$. Use a Lagrange multiplier $\lambda$.
Stationary equations are
\begin{equation}
RXCe-\lambda Re=e'Ce\times Ra.
\end{equation}
Premultiply by $e'$ gives
\begin{equation}
\hat\lambda=\frac{e'RXCe}{e'Re},
\end{equation}
and thus 
\begin{equation}
e'Ce\times\hat a=XCe-e\frac{e'RXCe}{e'Re},
\end{equation}
or
\begin{equation}
\hat a=\left(I-\frac{ee'R}{e'Re}\right)X\left(\frac{Ce}{e'Ce}\right).
\end{equation}

It is convenient to define the projectors
\begin{align}
P:=\frac{ee'R}{e'Re},\\
Q:=\frac{Cee'}{e'Ce}.
\end{align}
The projection on $\mathbb{L}_a$ is 
\begin{equation}
\hat ae'=(I-P)XQ.
\end{equation}

Projecting on $\mathbb{L}_b$ is similar, by symmetry. Thus
\begin{equation}
e\hat b'=PX(I-Q).
\end{equation}
In addition, from \eqref{eq-mu}, 
\begin{equation}
\hat\mu ee'=PXQ.
\end{equation}
It follows that if we define
\begin{equation}
\hat H:=X-\hat\mu ee'-\hat ae'-e\hat b'\label{eq-hdef}
\end{equation}
then
\begin{equation}
\hat H=(I-P)X(I-Q).
\end{equation}

Define $\mathbb{L}_h$ as the subspace of all matrices of this form, or equivalently the matrices with $HCe=0$ and $e'RH=0$.

Because $(I-Q)Ce=0$ it follows that
\begin{subequations}
\begin{equation}
\text{tr}\ R\hat HC(\hat\mu ee')=0
\end{equation}
and 
\begin{equation}
\text{tr}\ R\hat HC(e\hat b')=0
\end{equation}
and
\begin{equation}
\text{tr}\ R\hat ae'C\hat H'=0
\end{equation}
\end{subequations}
Thus $\mathbb{L}_h$ is also orthogonal in the Kronecker inner product to the other three subspaces. It follows that in the Kronecker norm
\begin{equation}
\|X\|^2=\|PXQ\|^2+\|(I-P)XQ\|^2+\|PX(I-Q)\|^2+\|(I-P)X(I-Q)\|^2,
\end{equation}
and thus we not only have a decomposition of the data $X$, but also of its Kronecker sum of squares.

It also follows that if $Y=\mu ee'+ae'+eb'+H$, with $a'Re=b'Ce=0$, then
\begin{multline}
\|X-Y\|^2=\|PXQ-\mu ee'\|^2+\|(I-P)XQ-ae'\|^2+\\\|PX(I-Q)-eb'\|^2+\|(I-P)X(I-Q)-H\|^2
\end{multline}
THus if we are fitting $Y$ to $X$ with seperable restrictions on $a$, $b$, and
$H$ (in addition to $a'Re=b'Ce=0$) then we can minmize each of the four 
components separately.

We have defined $H$ in \eqref{eq-hdef} as the residual after projecting
on the three orthogonal spaces. The orthogonality of the residual space
is YAPT (Yet Another Pythogorean Theorem). Alternatively, we could
have defined $\mathbb{L}_h$ as the space of $n\times m$ matrices
satisfying $HCe=0$ and $H'Re=0$ and project on this space. This will
produce the same $\hat H$.

The developments in this paper can be generalized in several ways. In the first
place the same reasoning applies to multiway arrays. If the array has $p$
"dimensions" then there are $2^p$ components. Secondly, 
vectors $e_n$ and $e_m$ in the formulas can be replaced by any vectors $u$ and $v$ with $n$ and $m$ elements. In fact, with some additional notation, they can
be replaced by matrices $U$ and $V$. And finally the results apply to
any finite sequence of orthogonal subspaces in any (finite-dimensional) inner
product space.



\sectionbreak


# Example

```{r}
set.seed(12345)
ymat <- matrix(rnorm(12), 4, 3)
rmat <- crossprod(matrix(rnorm(40), 10, 4)) / 10
cmat <- crossprod(matrix(rnorm(30), 10, 3)) / 10
```
```{r echo = FALSE}
print("ymat")
print(ymat)
print("cmat")
print(cmat)
print("rmat")
print(rmat)
```

\sectionbreak

We now apply the decompose() function from the code section.

```{r echo}
h<-decompose(ymat, rmat, cmat)
```
The four components are:
```{r echo = FALSE}
print(h$dd)
```

\sectionbreak

The Kronecker inner products of the four components are:
```{r echo = FALSE}
print(h$ip)
```
The trace of the inner product matrix and the squared norm of the data:
```{r echo = FALSE}
print(h$ssq)
```

\sectionbreak

# Code

```{r eval = FALSE}
set.seed(12345)
ymat <- matrix(rnorm(12), 4, 3)
rmat <- crossprod(matrix(rnorm(40), 10, 4)) / 10
cmat <- crossprod(matrix(rnorm(30), 10, 3)) / 10

decompose <- function(xmat, rmat, cmat) {
  n <- nrow(xmat)
  m <- ncol(xmat)
  pmat <- matrix(colSums(rmat), n, n, byrow = TRUE) / sum(rmat)
  qmat <- matrix(rowSums(cmat), m, m) / sum(cmat)
  px <- pmat %*% xmat
  xq <- xmat %*% qmat
  dd <- rep(list(0),4) 
  dd[[1]] <- px %*% qmat
  dd[[2]] <- px - dd[[1]]
  dd[[3]] <- xq - dd[[1]]
  dd[[4]] <- xmat - dd[[1]] - dd[[2]] - dd[[3]]
  ip <- matrix(0, 4, 4)
  for (i in 1:4) {
    for (j in 1:4) {
      ip[i, j] <- kroneckerIP(dd[[i]], dd[[j]], rmat, cmat)
    }
  }
  ssq <- c(kroneckerIP(xmat, xmat, rmat, cmat), sum(diag(ip)))
  return(list(dd = dd, ip = ip, ssq = ssq))
}

kroneckerIP <- function(xmat, ymat, rmat, cmat) {
  return(sum(ymat * (rmat %*% xmat %*% cmat)))
}
```
